---
title: "Lecture 16: Surveys"
author: "Brian J. Smith"
date: "2 April 2026"
format: 
  revealjs:
    width: 1400
    height: 900
    margin: 0.1
    theme: simple
    link-external-icon: true
    mermaid: 
      theme: forest
editor: source
---


```{r, setup}
#| include: false

library(ggplot2)
library(dplyr)
library(patchwork)
library(tidyr)

```

## Surveys

### Outline

- Visualizing Survey Results
- Graphics in Surveys
  - HTML Crash Course
  - Qualtrics
  - Google Forms
  
# Visualizing Survey Results

## Visualizing Survey Results

- Likert Scale
  - [Wikipedia](https://en.wikipedia.org/wiki/Likert_scale#Visual_presentation_of_Likert-type_data){target="_blank"}
    - [Robbins & Heiberger (2011)](http://www.asasrms.org/Proceedings/y2011/Files/300784_64164.pdf){target="_blank"}
    - [Heiberger & Robbins (2014)](https://doi.org/10.18637/jss.v057.i05){target="_blank"}
      - [`HH` R Package](https://cran.r-project.org/package=HH){target="_blank"}
  - [Daydreaming Numbers](https://daydreamingnumbers.com/4-ways-to-visualize-likert-scales/){target="_blank"}
  
```{r}
#| echo: true
#| eval: false

# Check this out:
library(HH)
demo("likert-paper")
```
  
## Visualizing Survey Results

- Fancy Stuff
  - [Flourish](https://flourish.studio/blog/visualizing-survey-data/){target="_blank"}

## Visualizing Survey Results

- Lots of Excel-based Advice
  - [SurveyMoneky](https://www.surveymonkey.com/learn/survey-best-practices/basic-chart-types-for-reports/){target="_blank"}
  - [Pointerpro](https://pointerpro.com/blog/survey-results-presentation/){target="_blank"}

# Graphics in Surveys

## Graphics in Surveys

### HTML Crash Course

- [HTML Cheat Sheet](https://syntaxsimplified.com/cheatsheet/HTML/hypertext_markup_language.html){target="_blank"}
- [HTML Cheat Sheet](https://web.stanford.edu/group/csp/cs21/htmlcheatsheet.pdf){target="_blank"}
- [HTML Cheat Sheet](https://htmlcheatsheet.com/){target="_blank"}

## Graphics in Surveys

### Qualtrics

- [How to access Qualtrics?](https://support.uidaho.edu/TDClient/40/Portal/KB/ArticleDet?ID=63){target="_blank"}
- [U of I Knowledge Base](https://support.uidaho.edu/TDClient/40/Portal/KB/?CategoryID=412){target="_blank"}
- [U of I Qualtrics](https://surveys.uidaho.edu){target="_blank"}

## Graphics in Surveys

### Google Forms

- [Google Forms](https://docs.google.com/forms){target="_blank"}
- [How to use Google Forms](https://support.google.com/docs/answer/6281888){target="_blank"}

## Graphics in Surveys

- [Qualtrics Survey](https://uidaho.co1.qualtrics.com/jfe/form/SV_4IaeTwJqxhiiQcK){target="_blank"}
- [Google Forms Survey](https://docs.google.com/forms/d/e/1FAIpQLSfELjskX-PzZZpGlfnSqO2MquH5ChOLtw529eQDytRmR6LO_g/viewform?usp=publish-editor){target="_blank"}
- [Code for Graphics](){target="_blank"}

# Questions?

</br>

</br>


:::{style="font-size: 50%"}
[BCB5200 Home](https://bsmity13.github.io/BCB5200/)
:::
