---
title: "Literate Programming"
format:
  html: 
    code-fold: true
    toc: true
---

# Heading 1
  
Writing some information, include some text that is **bold**.  

## Heading 2
  
Some more information, including some that is *italicized*.

## Heading 2
  
Again

# Tables

| Column 1 | Column 2 | Column 3 |
|----------|----------|----------|
| Row 1    | Row 1    | Row 1    |
| Row 2    | Row 2    | Row 2    |
| Row 3    | Row 3    | Row 3    |

# R Code

## Example 1

First, we can include code-formatted font in our narrative. So for example, if we wanted to say we were using the package `ggplot2` in our Methods, we can enclose it in "backticks" (\`).

## Example 2

Second, we can actually execute some code inline. For example, we might want to know what 1 + 1 equals: `{r} print(1+1)`.

## Example 3

Third, you might want a whole section of code (a "code chunk").

```{r ex3}
#| echo: true
# A bunch of R code to execute
# Attach sample data
data("mtcars")
# Print out the first 6 rows
head(mtcars)
```

## Example 4

Fourth, you might want to display the output without displaying the whole chunk of code.

```{r ex4}
#| echo: false

# A bunch of R code to execute
# Attach sample data
data("mtcars")
# Print out the first 6 rows
head(mtcars)
```

## Example 5

Since our output here was a table, we might want some formatting.

```{r}
#| echo: false

# A bunch of R code to execute
# Attach sample data
data("mtcars")
# Print out the first 6 rows
knitr::kable(head(mtcars))
```

## Example 6

We can include a figure in our output

```{r}
#| echo: false
#| warning: false
#| fig-cap: "Relationship between number of cylinders and gas mileage."
#| fig-width: 3
#| fig-asp: 0.67
#| fig-align: "left"

library(ggplot2)

ggplot(mtcars, aes(x = cyl, y = mpg)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_bw()


```